Solving Collective Spin Hamiltonians with Decoherence

### 2018-09-15 soc_squeezing/cumulant_expansions.tex
changed convention for sum limits in expectation value w.r.t. X
indexed variables to clarify their dependencies
fixed small error and removed extra whitespace

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{\ell,m,n} h_{\ell mn} S_+^\ell S_\z^m S_-^n,
  \label{eq:H_general}
\end{align}
where Hermiticity requires that $h_{\ell mn}=h_{nm\ell}^*$.  The
Hamiltonian for one-axis twisting with a transverse field, for
example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_\z^2 + \Omega S_\x
  = \chi S_\z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{\z,\y}$ and
$\p{\y,\x}$ are
\begin{align}
  H_{\t{TAT}}^{\z,\y}
  \propto S_\z^2 - S_\y^2
  = S_\z^2 + \f14 S_+^2 + \f14 S_-^2
  - \f12 S_+ S_- + \f12 S_\z,
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{\y,\x}
  \propto S_\y^2 - S_\x^2
  = -\f12 S_+^2 - \f12 S_-^2.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_\z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = i \sum_{p,q,r} h_{pqr}
  \sp{S_+^p S_\z^q S_-^r, S_+^\ell S_\z^m S_-^n}_-
  + \sum_\J \gamma_\J \D\p{\J;\O}
  \equiv \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_\z^q S_-^r,
  \label{eq:time_derivative}
\end{align}
where commutator and decoherence terms necessary for computing the
matrix elements $T^{\ell m n}_{pqr}$ of the time derivative operator
$d/dt$ are worked out Appendices \ref{sec:comm_general},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective}.

@@

The truncation strategy in \eqref{eq:time_derivative_truncation} has a
few limitations:
\begin{enumerate*}[label=(\roman*)]
\item simulating a system of differential equations for a large number
  of operators can be very time-consuming,
\item the weight function $w$ must be chosen carefully, as the optimal
  weight function is generally system-dependent,
\item simulation results can only be trusted up to the time at which
  the initial value of operators $S_+^p S_\z^r S_-^q$ with weight
  $w\p{p,q,r}\ge W$ have a non-negligible contribution to operators of
  interest.
  \label{pt:limitation}
\end{enumerate*}
While the first two of these limitations might in principle be avoided
by alternate means of tracking operators, the last limitation
necessarily applies in some form to any method tracking only a subset
of {\it all} operators.  Embracing the unavoidable, we therefore try
to devise an alternate truncation strategy which is built around
\ref{pt:limitation}.

@@

Initial conditions for all collective spin operators can be determined
from the expansions
\begin{align}
  S_\z = \sum_{k=-S}^S k \op{k},
  &&
  S_\mu = \sum_{k=-S+\delta_{\mu,-1}}^{S-\delta_{\mu,1}}
  g_\mu\p{S,k} \op{k+\mu}{k},
\end{align}
with
\begin{align}
  g_\mu\p{S,k} \equiv \sqrt{\p{S-\mu k}\p{S+\mu k+1}},
\end{align}
and the initial states
\begin{align}
  \ket{\pm\Z} \equiv \ket{\pm S},
  &&
  \ket{\X} \equiv 2^{-N/2} \sum_{k=-S}^S
  { N \choose S+k }^{1/2} \ket{k},
\end{align}
using which we can compute, for $\mu,\nu\in\set{+1,-1}$,
$\bmu\equiv-\mu$,
\begin{align}
  \bk{\nu\Z|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\nu\Z}
  = \delta_{\ell n} \times
  \begin{cases}
     \p{S-n}^m \f{N! n!}{\p{N-n}!} & \mu = \nu, \\
     \delta_{n,0} \p{-S}^m & \mu \ne \nu,
  \end{cases}
\end{align}
and
\begin{align}
  \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}
  &=  \f{\mu^m}{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}} k^m
  \sp{{ N \choose S+k+\mu\ell } { N \choose S+k+\mu n }}^{1/2}
  \notag \\
  &\qquad\qquad\qquad\qquad
  \times \prod_{p=0}^{\ell-1} g_\mu\p{S,k+\mu p}
  \prod_{q=0}^{n-1} g_\bmu\p{S,k+\mu\sp{n-q}} \\
  &= \mu^m \f{N!}{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}} k^m
  \f{\p{S-\mu k}!}{\p{S+\mu k}!\p{S-\mu k-\ell}!\p{S-\mu k-n}!} \\
  &= \p{-1}^m \f{N!}{2^N} \sum_{k=-S+\max\set{\ell,n}}^S k^m
  \f{\p{S+k}!}{\p{S-k}!\p{S+k-\ell}!\p{S+k-n}!}.
\end{align}
Furthermore, using the fact that
\begin{align}
  e^{i\phi S_\z} S_\mu e^{-i\phi S_\z} = e^{i\mu\phi} S_\mu,
\end{align}
we can say
\begin{align}
  \bk{\phi|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\phi}
  = e^{i\mu\p{\ell-n}\phi} \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X},
\end{align}
where
\begin{align}
  \ket{\phi} \equiv e^{-i\phi S_\z}\ket{\X},
  &&
  \p{\cos\phi S_\x + \sin\phi S_\y} \ket\phi = S \ket\phi.
\end{align}
Alternatively, with the help of Mathematica we can find
\begin{align}
  \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}
  = \f{\p{N!}^2}{2^N\p{N-\ell}!\p{N-n}!}
  \sp{\p{-S}^m \alpha + \f{\p{N-\ell}\p{N-n}}{N} \beta},
\end{align}
where $\alpha$ and $\beta$ are defined in terms of the generalized
hypergeometric function $\F$ by
\begin{align}
  \alpha
  \equiv \F\p{\let\scriptstyle\textstyle
    \substack{\ell-N,~n-N \\ -N}~; -1},
\end{align}
and
\begin{align}
  \beta
  \equiv \sum_{p=0}^{m-1} \p{-S}^p { m \choose p }
  \tilde\F_{m-p-1}\p{\let\scriptstyle\textstyle
    \substack{1+\ell-N,~1+n-N \\ 1-N}~; -1},
\end{align}
for
\begin{align}
  \tilde\F_q\p{\let\scriptstyle\textstyle
    \substack{a,~b\\c}~; z}
  \equiv \F\p{\let\scriptstyle\textstyle
    \substack{2,2,\cdots,2,a,b \\ 1,1,\cdots,1,c}~; z}
\end{align}
with $q$ repetitions of 2 and 1.

>>>>

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{\ell,m,n} h_{\ell mn} S_+^\ell S_\z^m S_-^n,
  \label{eq:H_general}
\end{align}
where Hermiticity requires that $h_{\ell mn}=h_{nm\ell}^*$.  The
Hamiltonian for one-axis twisting with a transverse field, for
example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_\z^2 + \Omega S_\x
  = \chi S_\z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{\z,\y}$ and
$\p{\y,\x}$ are
\begin{align}
  H_{\t{TAT}}^{\z,\y}
  \propto S_\z^2 - S_\y^2
  = S_\z^2 + \f14 S_+^2 + \f14 S_-^2
  - \f12 S_+ S_- + \f12 S_\z,
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{\y,\x}
  \propto S_\y^2 - S_\x^2
  = -\f12 S_+^2 - \f12 S_-^2.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_\z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = i \sum_{p,q,r} h_{pqr}
  \sp{S_+^p S_\z^q S_-^r, S_+^\ell S_\z^m S_-^n}_-
  + \sum_\J \gamma_\J \D\p{\J; S_+^\ell S_-^m S_\z^n}
  \equiv \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_\z^q S_-^r,
  \label{eq:time_derivative}
\end{align}
where commutator and decoherence terms necessary for computing the
matrix elements $T^{\ell m n}_{pqr}$ of the time derivative operator
$d/dt$ are worked out Appendices \ref{sec:comm_general},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective}.

@@

The truncation strategy in \eqref{eq:time_derivative_truncation} has a
few limitations:
\begin{enumerate*}[label=(\roman*)]
\item simulating a system of differential equations for a large number
  of operators can be very time-consuming,
\item the weight function $w$ must be chosen carefully, as the optimal
  weight function is generally system-dependent,
\item simulation results can only be trusted up to the time at which
  the initial value of operators $S_+^p S_\z^r S_-^q$ with weight
  $w\p{p,q,r}\ge W$ have a non-negligible contribution to operators of
  interest.\label{pt:limitation}
\end{enumerate*}
While the first two of these limitations might in principle be avoided
by alternate means of tracking operators, the last limitation
necessarily applies in some form to any method tracking only a subset
of {\it all} operators.  Embracing the unavoidable, we therefore try
to devise an alternate truncation strategy which is built around
\ref{pt:limitation}.

@@

Initial conditions for all collective spin operators can be determined
from the expansions
\begin{align}
  S_\z = \sum_{k=-S}^S k \op{k},
  &&
  S_\mu = \sum_{k=-S+\delta_{\mu,-1}}^{S-\delta_{\mu,1}}
  g_\mu\p{S,k} \op{k+\mu}{k},
\end{align}
with
\begin{align}
  g_\mu\p{S,k} \equiv \sqrt{\p{S-\mu k}\p{S+\mu k+1}},
\end{align}
and the initial states
\begin{align}
  \ket{\pm\Z} \equiv \ket{\pm S},
  &&
  \ket{\X} \equiv 2^{-N/2} \sum_{k=-S}^S
  { N \choose S+k }^{1/2} \ket{k},
\end{align}
using which we can compute, for $\mu,\nu\in\set{+1,-1}$,
$\bmu\equiv-\mu$,
\begin{align}
  \bk{\nu\Z|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\nu\Z}
  = \delta_{\ell n} \times
  \begin{cases}
     \p{S-n}^m \f{N! n!}{\p{N-n}!} & \mu = \nu, \\
     \delta_{n,0} \p{-S}^m & \mu \ne \nu,
  \end{cases}
\end{align}
and
\begin{align}
  \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}
  &=  \f{\mu^m}{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}} k^m
  \sp{{ N \choose S+k+\mu\ell } { N \choose S+k+\mu n }}^{1/2}
  \notag \\
  &\qquad\qquad\qquad\qquad
  \times \prod_{p=0}^{\ell-1} g_\mu\p{S,k+\mu p}
  \prod_{q=0}^{n-1} g_\bmu\p{S,k+\mu\sp{n-q}} \\
  &= \mu^m \f{N!}{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}}
  \f{k^m \p{S-\mu k}!}
  {\p{S+\mu k}!\p{S-\mu k-\ell}!\p{S-\mu k-n}!} \\
  &= \f{N!}{2^N} \sum_{k=-S}^{S-\max\set{\ell,n}}
  \f{k^m \p{S-k}!}{\p{S+k}!\p{S-k-\ell}!\p{S-k-n}!}.
\end{align}
Furthermore, using the fact that
\begin{align}
  e^{i\phi S_\z} S_\mu e^{-i\phi S_\z} = e^{i\mu\phi} S_\mu,
\end{align}
we can say
\begin{align}
  \bk{\phi|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\phi}
  = e^{i\mu\p{\ell-n}\phi} \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X},
\end{align}
where
\begin{align}
  \ket{\phi} \equiv e^{-i\phi S_\z}\ket{\X},
  &&
  \p{\cos\phi S_\x + \sin\phi S_\y} \ket\phi = S \ket\phi.
\end{align}
Alternatively, with the help of Mathematica we can find
\begin{align}
  \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}
  = \f{\p{N!}^2}{2^N\p{N-\ell}!\p{N-n}!}
  \sp{\p{-S}^m \alpha_{N\ell n}
    + \f{\p{N-\ell}\p{N-n}}{N} \beta_{N\ell mn}},
\end{align}
where $\alpha$ and $\beta$ are defined in terms of the generalized
hypergeometric function $\F$ by
\begin{align}
  \alpha_{N\ell n}
  \equiv \F\p{\let\scriptstyle\textstyle
    \substack{\ell-N,~n-N \\ -N}~; -1},
\end{align}
and
\begin{align}
  \beta_{N\ell mn}
  \equiv \sum_{p=0}^{m-1} \p{-S}^p { m \choose p }
  \tilde\F_{m-p-1}\p{\let\scriptstyle\textstyle
    \substack{1+\ell-N,~1+n-N \\ 1-N}~; -1},
\end{align}
for
\begin{align}
  \tilde\F_q\p{\let\scriptstyle\textstyle
    \substack{a,~b\\c}~; z}
  \equiv \F\p{\let\scriptstyle\textstyle
    \substack{2,2,\cdots,2,a,b \\ 1,1,\cdots,1,c}~; z}
\end{align}
with $q$ repetitions of 2 and 1.
